---
title: "Untitled"
---

```{r}
#| echo: false
suppressPackageStartupMessages({
  library(tidyverse)
  library(scales)
  library(patchwork)
})
```

## The logic of hypothesis testing

Hypothesis testing is effectively a signal-to-noise problem. When a difference (signal) is observed, one must decide how probable it is that this difference is meaningful versus merely the result of random variation (noise).

Imagine a company launches a new chocolate product. They claim the mass of their product is 50g $\pm$ 1g. What that means is that the average mass of their chocolate This means they claim the average mass is 50g, but due to machine imperfection, individual bars fluctuate. Some might be 49.2g, others 50.8g. This fluctuation is considered noise in our example and in statistics is known as standard deviation ($\sigma$).

Now, suppose we buy a chocolate bar and measure its mass at 51.7g.The difference ($51.7 - 50 = 1.7$g) is the signal we observe.The critical question is: How probable is it that this signal is simply due to variability in chocolate mass (noise)? To quantify this probability, we introduce the Probability Density Function (PDF). This curve serves as a "map" of the noise, showing us exactly how likely different fluctuations are.

```{r}
# Parameters
suppressPackageStartupMessages({
  library(tidyverse)
  library(scales)
})

mu <- 50
sigma <- 1

# Create data for the curve
plot_data <- tibble(
  x = seq(mu - 4 * sigma, mu + 4 * sigma, length.out = 1000),
  y = dnorm(x, mean = mu, sd = sigma)
)

# Custom Label Function: Stacks IQ on top of Sigma
dual_label <- function(x) {
  # Calculate Z-score (Sigma)
  z <- (x - mu) / sigma
  
  # Format the Sigma string (e.g., "+1σ")
  sigma_str <- paste0(ifelse(z > 0, "+", ""), z, "\u03C3") # \u03C3 is sigma symbol
  sigma_str[z == 0] <- "\u03BC" # Use mu symbol for 0
  
  # Return: "IQ \n Sigma"
  paste0(x, "\n", sigma_str)
}

ggplot(plot_data, aes(x = x, y = y)) +
  geom_area(
    data = filter(plot_data, x >= mu - 3 * sigma & x <= mu + 3 * sigma),
    fill = "#DEEBF7"
  ) +
  geom_area(
    data = filter(plot_data, x >= mu - 2 * sigma & x <= mu + 2 * sigma),
    fill = "#9ECAE1"
  ) +
  geom_area(
    data = filter(plot_data, x >= mu - 1 * sigma & x <= mu + 1 * sigma),
    fill = "#4292C6"
  ) +
  geom_line(linewidth = 1) +
  scale_x_continuous(
    name = "IQ",
    breaks = seq(46, 54, 1),
    labels = dual_label # Apply our custom stacker
  ) +
  scale_y_continuous(
    "Probability density",
    expand = expansion(mult = c(0, 0.01))
  ) +
  
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 11, lineheight = 1.2),
    axis.title.x = element_text(margin = margin(t = 15)),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank()
  )

```

The **Z-test** is the foundational model for this process. While rarely used in practice (because it requires knowing the population variance), it is pretty good to understand how we map raw data onto probability distributions.

### The scenario

Imagine we measure the IQ of 30 students who competed in a Math Olympiad.

-   **Observed sample mean (**$\bar{x}$): 118.5

-   **Sample size (**$n$): 30

We assume the parameters for the general human population are known facts:

-   **Population mean (**$\mu$): 110

-   **Population SD (**$\sigma$): 15

**Research question:** Is there sufficient evidence to conclude that these students differ significantly from the general population?

### Assumptions

-   **Continuous data:** The dependent variable is continuous.

-   **Independence:** Observations are independent.

-   **Normality:** The sampling distribution is normal (satisfied by CLT as $n=30$).

-   **Known variance:** The population $\sigma$ is known.

### 1. Visualizing the mechanism

To test the hypothesis, we must compare our sample against the "Null distribution."

Crucially, we do not test against the distribution of *individuals* (Population). We test against the distribution of *means* (Sampling Distribution). The Z-test relies on the fact that the Sampling Distribution is much narrower than the Population, making it sensitive to deviations.

```{r}
#| label: mechanism-plot
#| fig-height: 6
#| fig-width: 8
#| echo: false

# Parameters
mu_pop    <- 110
sigma_pop <- 15
x_bar     <- 118.5
n         <- 30
sem       <- sigma_pop / sqrt(n)
common_xlim <- c(60, 160)

# PLOT A: POPULATION (Individuals)
p1 <- ggplot(data = tibble(x = common_xlim), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = mu_pop, sd = sigma_pop)) +
  
  # Shade 1 SD spread
  stat_function(fun = dnorm, args = list(mean = mu_pop, sd = sigma_pop),
                xlim = c(mu_pop - sigma_pop, mu_pop + sigma_pop),
                geom = "area", aes(fill = "1 SD spread"), alpha = 0.5) +
  
  geom_vline(xintercept = x_bar, color = "blue", linewidth = 0.8) +
  
  scale_fill_manual(name = NULL, values = c("1 SD spread" = "gray80")) +
  theme_classic() +
  labs(title = "1. Population distribution (Individuals)",
       subtitle = "Wide variance (Sigma = 15). An individual IQ of 118.5 is common.",
       y = "Density", x = NULL) +
  coord_cartesian(xlim = common_xlim)

# PLOT B: SAMPLING DISTRIBUTION (Means)
p2 <- ggplot(data = tibble(x = common_xlim), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = mu_pop, sd = sem)) +
  
  # Shade P-value region
  stat_function(fun = dnorm, args = list(mean = mu_pop, sd = sem),
                xlim = c(x_bar, 160),
                geom = "area", aes(fill = "P-value region"), alpha = 0.6) +
  
  geom_vline(xintercept = x_bar, color = "blue", linewidth = 0.8) +
  
  scale_fill_manual(name = NULL, values = c("P-value region" = "red")) +
  theme_classic() +
  labs(title = "2. Sampling distribution (Means of n=30)",
       subtitle = "Narrow variance (SEM = 2.74). A group average of 118.5 is rare.",
       y = "Density", x = "IQ score") +
  coord_cartesian(xlim = common_xlim)

# Combine plots
p1 / p2
```

### 2. The calculation

The Z-test quantifies the visual difference above. It calculates the distance between the observed mean and the expected mean in units of **Standard Error**.

**Step A: Calculate standard error (SEM)**

The spread of the sampling distribution (Plot 2) is determined by the population $\sigma$ and sample size $n$.

\$\$SEM = \\frac{\\sigma}{\\sqrt{n}} = \\frac{15}{\\sqrt{30}} \\approx 2.74\$\$

**Step B: Calculate Z-score**

We standardize the difference. This maps our specific experiment (IQ) onto the universal Standard normal distribution.

\$\$Z = \\frac{\\bar{x} - \\mu}{SEM} = \\frac{118.5 - 110}{2.74} \\approx 3.10\$\$
