---
author: Amel Zulji
title: "Sampling and central limit theorem"
description: "Central limit theorem and its implications"
categories:
  - statistics
execute: 
  echo: false
---

```{r}
#| echo: false

set.seed(12345)

suppressPackageStartupMessages({
  library(tidyverse)
  library(scales)
  library(patchwork)
  library(eurostat)
})

source("helpers/geom_analysis.R")

theme_set(
  theme_bw() + 
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.tag = element_text(size = 16, face = "bold")
      )
)

```

# Sampling fundamentals

To understand sampling, lets first define population, its distribution, and its parameters.

### The population

A population is the complete set of all possible observations for a specific variable of interest. It represents the ground truth for the variable of interest. For example, the height (variable of interest) of all male humans (all possible observations) would represent a population.

### 2. Probability density function and population distribution

If we consider the heights of all adult males, we notice that different heights occur with different frequencies. For example, a randomly selected male is much more likely to be 180 cm tall than 220 cm.

This means the population is "denser" around 180 cm - there are significantly more individuals with a height around 180 cm than eg around 220 cm.

The Probability Density Function (PDF) mathematically defines this density. Plotting the density as a function of height provides a continuous representation of the population's distribution, showing us instantly where the "mass" of the population lies. (@fig-heigh-and-income)

```{r}
#| label: fig-heigh-and-income
#| warning: false
#| fig-cap: Distribution of human male heights (a) and Personal income (b)

N <- 10000
pop1_name <- "Human male height"
pop1_unit <- "cm"
pop1_mu <- 175
pop1_sigma <- 10
pop1 <- rnorm(N, pop1_mu, pop1_sigma) |> enframe(name = NULL)
pop1_p <- pop1 |>
  ggplot(aes(x = value)) +
  geom_density() +
  scale_x_continuous(name = str_glue("{pop1_name} / {pop1_unit}"))

pop2_stat <- get_eurostat("ilc_di03", time_format = "date") |>
  filter(
    geo == "DE",
    age == "TOTAL",
    sex == "T",
    TIME_PERIOD == "2024-01-01",
    unit == "EUR"
  ) |>
  select(indic_il, values) |>
  deframe()

pop2_name <- "Personal income"
pop2_unit <- "€"
pop2_mu <- pop2_stat[["MEI_E"]]
pop2_median <- pop2_stat[["MED_E"]]
pop2 <- rlnorm(N, log(pop2_median), sqrt(2 * log(pop2_mu / pop2_median))) |> enframe()
pop2_p <- pop2 |>
  ggplot(aes(x = value)) +
  geom_density() +
  scale_x_continuous(name = str_glue("{pop2_name} / {pop2_unit}")) + 
  scale_y_continuous(labels = scales::label_number())

pop1_p

# pop1_p + pop2_p + plot_annotation(tag_levels = "a")

```

We see in the @fig-heigh-and-income that the density is heighst

```{r}

pop1_p +
  geom_function(
    fun = \(x) dnorm(x = x, mean = pop1_mu, sd = pop1_sigma),
    mapping = aes(linetype = "Nomral distribution"),
    color = "red"
  ) +
  labs(linetype = NULL)

```
