---
title: "Testing drafts"
draft: true
---

::: callout-warning
-   Currently it works by saving the list directly in excel so that you get the list names as sheet. Then you load saved excel as wb object to apply styling. Idealy you should find a way to include dataframes programatically like this\
    `wb <- wb_workbook() %>% wb_add_worksheet(sheet = "test1") %>% wb_add_data(x = mtcars)`if this doesnt work check to do it with `imap` where `y` will be the name

-   currently it also lacks programatic way to check number of columns to apply styling to.

Perhaps `imap` would solve both problems
:::

When drafting a figure you collect data which should be reported as supplementary tables in publication. Some journals require to report supplementary table in .xlsx format. The code bellow demonstrates how to do that programatically with R. In particular, it shows how to save a named list of dataframes as in .xlsx format where each dataframe will be saved as separate sheet named according to its name in list.

```{r}
renv::use(lockfile = "renv.lock", verbose = FALSE)
```

Create a named list

```{r}
library(tidyverse)
library(openxlsx2)
tmp_list <- list(df1 = mtcars, df2 = mtcars, df3 = mtcars)
map(tmp_list, head)
```

Save files in xlsx.

```{r}
tmp_path <- tempfile(fileext = ".xlsx")
write_xlsx(tmp_list, tmp_path)
```

::: callout-note
This excel file file have sheets named df1,df2,df3
:::

Apply styling to excel file programtically

```{r}
wb_exc <- wb_load(tmp_path)
map(wb_exc$sheet_names, \(x) {
  wb_exc <<-  wb_exc %>% wb_set_col_widths(cols = 1:12, widths = "auto")
})
wb_save(wb = wb_exc, file = tmp_path)
```
