---
author: Amel Zulji
title: "Central limit theorem"
description: "Central limit theorem and its implications"
categories:
  - statistics
execute: 
  echo: false
---

```{r}
#| echo: false

set.seed(12345)

suppressPackageStartupMessages({
  library(tidyverse)
  library(scales)
  library(patchwork)
  
})

# define new geom 
StatDensityShade <- ggproto(
  "StatDensityShade",
  Stat,
  compute_group = function(data, scales) {
    dens <- density(data$x)
    mu <- mean(data$x)
    sigma <- sd(data$x)

    df <- data.frame(x = dens$x, y = dens$y)
    df_shade <- df[df$x >= (mu - sigma) & df$x <= (mu + sigma), ]

    df_shade <- rbind(c(min(df_shade$x), 0), df_shade, c(max(df_shade$x), 0))
    return(df_shade)
  },
  required_aes = c("x")
)

StatDensityMean <- ggproto(
  "StatDensityMean",
  Stat,
  compute_group = function(data, scales) {
    dens <- density(data$x)
    mu <- mean(data$x)
    y_height <- approx(dens$x, dens$y, xout = mu)$y
    data.frame(x = mu, xend = mu, y = 0, yend = y_height)
  },
  required_aes = c("x")
)

geom_density_sigma <- function(
  fill_color = "skyblue",
  line_color = "red",
  fill_label = "mu %+-% sigma",
  line_label = "mu",
  line_type = "solid",
  alpha = 0.5,
  ...
) {
  list(
    geom_polygon(
      stat = StatDensityShade,
      aes(fill = fill_label),
      alpha = alpha
    ),

    geom_density(fill = NA, linewidth = 0.7, ...),

    geom_segment(
      stat = StatDensityMean,
      aes(color = line_label),
      linetype = line_type,
      linewidth = 0.7
    ),
    scale_fill_manual(
      name = "Spread",
      values = setNames(fill_color, fill_label),
      labels = scales::label_parse()
    ),
    scale_color_manual(
      name = "Centrality",
      values = setNames(line_color, line_label),
      labels = scales::label_parse()
    )
  )
}
```

## Introduction

To understand the Central Limit Theorem (CLT), the concepts of population, sampling, and sampling distributions must first be understood (@fig-pop-sam-samdist).

-   **Population** is defined as the set of all possible observations about a specific variable. A population is usually characterized by statistical parameters such as centrality (mean $\mu$, median, mode) and spread (variance $\sigma^2$, standard deviation $\sigma$). The population distribution describes how the values are spread across the population. The most common distribution is the normal distribution, which is fully defined by its mean (centrality) and standard deviation (spread). In statistics, a normally distributed population is denoted as $N(\mu, \sigma)$.

-   **Sample** represents a subset of the population. In a simple random sample, observations are taken in such a way that every member of the population has an equal chance of being selected. Like a population, a sample is defined by centrality and spread. However, to distinguish them from population parameters, different notation (Latin letters) is used for sample statistics.

-   **Sampling distribution** represents the distribution of a statistical parameter (such as the mean) calculated from many different samples drawn from the same population.

```{r}
#| warning: false
#| label: fig-pop-sam-samdist
#| fig-cap: Relationship between population (a), sample (b), and sampling distribution (c).
#| fig-align: center
#| fig-height: 7
#| fig-width: 5

pop <- rnorm(10000, 100, 15)
xmin <- mean(pop) - 3 * sd(pop)
xmax <- mean(pop) + 3 * sd(pop)

p1 <- pop |>
  enframe() |>
  ggplot(aes(x = value)) +
  geom_density_sigma(fill_color = "green") +
  xlim(xmin, xmax) + 
  ggtitle("Population distirbution") +
  theme_bw()

p2 <- sample(pop, 20) |>
  enframe() |>
  ggplot(aes(x = value)) +
  geom_density_sigma(fill_label = "bar(x) %+-% s", fill_color = "orange", line_label = "bar(x)", line_type = "1111") +
  xlim(xmin, xmax) + 
  ggtitle("Sample distirbution") +
  theme_bw()

p3 <- map_dbl(1:1000, \(x) sample(pop, 20) |> mean()) |>
  enframe() |>
  ggplot(aes(x = value)) +
  geom_density_sigma(fill_label = "mu[bar(x)] %+-% sigma[bar(x)]", fill_color = "lightblue", line_label = "mu[bar(x)]",line_type = "2222") +
  xlim(xmin, xmax) + 
  ggtitle("Sampling distirbution of the mean") +
  theme_bw()

p1 / p2 / p3 + plot_annotation(tag_levels = "a") & theme(plot.tag = element_text(size = 15, face = "bold"))

```

## Central Limit Theorem

The **CLT** states that the sampling distribution of the mean approaches a normal distribution as the sample size increases, regardless of the shape of the underlying population distribution from which the samples are drawn, provided that the population has a finite mean and variance.

To demonstrate this, three populations are simulated. Repetitive sampling (1 000 repetitions) is performed for various sample sizes ($n$), and the changes in the distribution of the sampling mean are observed (@fig-sampling-dist)

```{r}
#| warning: false
#| code-fold: true
#| fig-width: 7
#| fig-height: 7
#| label: fig-sampling-dist
#| fig-cap: Visualizing the central limit theorem. The top row displays distribution of normal, uniform, and bimodal population. The rows below display the sampling distribution of the mean for increasing sample sizes (n = 3, 5, 10, 20).

N <- 100000
n <- c(3, 5, 10, 20)

pop_list <- list(
  Normal = rnorm(N, 0.5, 0.1),
  Uniform = runif(N, 0, 1),
  Random = c(rnorm(N/2, mean = 0, sd = 1), rnorm(N/2, mean = 3.5, sd = 1))
)

pop_dist <- imap(pop_list, \(pop, pop_name) {
  lim <- c(xmin = mean(pop) - 3 * sd(pop), xmax = mean(pop) + 3 * sd(pop))
  p1 <- pop |>
    enframe() |>
    ggplot(aes(x = value)) +
    geom_density_sigma(fill_color = "green") +
    xlim(lim["xmin"], lim["xmax"]) +
    ggtitle(pop_name) +
    theme_bw()
  p2 <- map(n, \(x) {
    facet_lab <- paste("n =", n)
    replicate(n = 1000, expr = sample(pop, x) |> mean()) |>
    enframe() |>
    ggplot(aes(x = value)) +
    geom_density_sigma(
      fill_label = "sigma[bar(x)]",
      fill_color = "lightblue",
      line_label = "mu[bar(x)]",
      line_type = "2222"
    ) +
    xlim(lim["xmin"], lim["xmax"]) +
    theme_bw()
  })
  
  wrap_plots(c(list(p1), p2), ncol = 1)
})

wrap_plots(pop_dist, nrow = 1, guides = "collect") &
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

## Implications

Beyond the convergence of shape, the CLT provides precise predictions regarding the parameters of the sampling distribution. Specifically, it relates the mean and standard deviation of the sample means to the underlying population parameters.

The center of the sampling distribution ($\mu_{\bar{x}}$) is equal to the mean of the original population ($\mu$):

$\mu_{\bar{x}} = \mu$

The variability of the sampling distribution, $\sigma_{\bar{x}}$ (known as the Standard Error of the Mean or S.E.M.), is directly related to the population standard deviation ($\sigma$) and the sample size ($n$). As observed in the visual simulation above, the spread decreases as $n$ increases. This relationship is defined by the inverse-square law:

$\sigma_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$

To validate these theoretical properties, a second simulation was conducted. Three independent samples were drawn from a normally distributed population. Cumulative statistics (Mean, SD, and SEM) were calculated as data points were added one by one (from $n=2$ to $n=100$). The empirical standard deviation of the sample means was then compared against the theoretical prediction ($\sigma / \sqrt{n}$) as illustrated in @fig-clt-implications

```{r}
#| label: fig-clt-implications
#| code-fold: true
#| fig-height: 7
#| fig-width: 7
#| fig-align: center
#| fig-cap: Convergence of sample statistics. The plots illustrate how cumulative statistics evolve as the sample size increases from 2 to 100.

n_size <- 100
n_samples <- 3
true_mu <- 0.5
true_sigma <- 0.1

df_cum <- expand_grid(sample_id = paste("Sample", 1:n_samples), n = 1:n_size) |>
  mutate(value = rnorm(n(), true_mu, true_sigma)) |>
  group_by(sample_id) |>
  mutate(
    Mean = cummean(value), # Cumulative mean
    SD = map_dbl(n, \(i) sd(value[1:i])), # cumulative SD
    SEM = SD / sqrt(n) # umulative SE
  ) |>
  filter(n > 1) |>
  ungroup()

# Reshape for plotting
plot_data_cum <- df_cum |>
  pivot_longer(
    cols = c(Mean, SD, SEM),
    names_to = "metric",
    values_to = "observed"
  ) |>
  mutate(
    theoretical = case_when(
      metric == "Mean" ~ true_mu,
      metric == "SD" ~ true_sigma,
      metric == "SEM" ~ true_sigma / sqrt(n)
    ),
    # Create Labels for facets
    metric_label = factor(
      metric,
      levels = c("Mean", "SD", "SEM"),
      labels = c(
        "Cumulative~Mean~(bar(x))",
        "Cumulative~SD~(s)",
        "Cumulative~SEM~(s/sqrt(n))"
      )
    )
  )

# Plot the convergence
ggplot(plot_data_cum, aes(x = n)) +
  geom_point(aes(y = observed, color = sample_id), size = 1.5, alpha = 0.6) +
  geom_line(
    aes(y = theoretical, linetype = "True Parameter"),
    linewidth = 1.2
  ) +

  facet_wrap(
    ~metric_label,
    ncol = 1,
    scales = "free_y",
    labeller = label_parsed
  ) +

  scale_color_brewer(palette = "Set1", name = NULL) +
  scale_linetype_manual(name = NULL, values = c("True Parameter" = "1111")) +

  theme_bw() +
  labs(
    y = NULL,
    x = "Sample Size (n)"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill = "gray90")
  )
```

Finally, an important practical question is how large the sample size $n$ must be for the **sampling distribution of the mean** to become approximately normal. To address this, four populations are simulated: a normal distribution, a mildly right-skewed distribution, a moderately right-skewed distribution, and an extremely right-skewed distribution.

From each population, samples are repeatedly drawn (1,000 repetitions) with sizes ranging from $n=2$ to $n=60$. For each $n$, the approach of the sampling distribution of the mean toward normality is examined. Normality is assessed using the **Shapiro–Wilk** $W$ statistic. Each result is color-coded based on the corresponding $p$-value: values greater than 0.05 (red) indicate no evidence against normality, whereas values below 0.05 (blue) indicate a deviation from normality (@fig-clt-normality).

```{r}
#| label: fig-clt-normality
#| warning: false
#| code-fold: true
#| fig-height: 6
#| fig-width: 9
#| fig-align: center
#| fig-cap: Convergence of sampling distribution of the mean to normality.

pop_list <- list(
  Normal = rnorm(N, mean = 0.5, sd = 0.3),
  "Mild-right-skew" = rbeta(N, shape1 = 1, shape2 = 2),
  "Moderate-right-skew" = rbeta(N, shape1 = 1, shape2 = 3),
  "Extreme-right-skew" = rbeta(N, shape1 = 1, shape2 = 6)
)


pop_dist <- imap(pop_list, \(pop, pop_name) {
  lim <- c(xmin = mean(pop) - 3 * sd(pop), xmax = mean(pop) + 3 * sd(pop))
  p1 <- pop |>
    enframe() |>
    ggplot(aes(x = value)) +
    geom_density_sigma(fill_color = "green") +
    # xlim(lim["xmin"], lim["xmax"]) +
    xlim(-0.2, 1.2) +
    ylim(0, 6) +
    ggtitle(pop_name) +
    theme_bw()
  
  results <- map(1:60 |> set_names(), \(n_size) {
    # Draw samples and compute means
    sample_means <- replicate(1000, mean(sample(pop, size = n_size)))
    
    # Run Shapiro Test
    st <- shapiro.test(sample_means)
    
    # Return multiple metrics
    tibble(
      n = n_size,
      p_value = st$p.value,
      W_stat  = st$statistic,
    ) |> 
      mutate(
        Normality = case_when(p_value > 0.05 ~ "Normal", T ~ "Not normal")
      )
  }) |> list_rbind()
  
  p2 <- results |> 
    ggplot(aes(x = n, y = W_stat)) + 
    geom_line(color = "blue") +
    geom_point(aes(color = Normality), size = 1.5) +
    geom_smooth(se = FALSE, ) +
    ylim(0.94,1) +
    scale_x_continuous(breaks = 1:6*10) +
    labs(y = "W statistic (shapiro test)") + 
    theme_bw()
  
  p1 / p2
  
})

fig <- wrap_plots(pop_dist, nrow = 1, guides = "collect", axes = "collect")
for (i in 2:length(pop_list)) {
  for (j in 1:2) {
    fig[[i]][[j]] <- fig[[i]][[j]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
  }
}

fig & theme(legend.position = "bottom")

```

It is observed that for a normally distributed population, the sampling distribution of the mean is already approximately normal even for very small sample sizes (e.g., $n=3$). As the underlying population departs from normality and becomes increasingly right-skewed, larger sample sizes are required. In these simulations, a sample size of approximately $n \approx 30$ is found to be sufficient for the sampling distribution of the mean to approach normality, even in the case of an extremely right-skewed population.
